Structural Stability of Random Dynamical Systems

### 2014-09-05 journal.tex
update text to match computations

\section{$2 \times 2$ systems}
We call a matrix stable when all its eigenvalues are negative.  For a
$2 \times 2$ matrix, this amounts to the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant.

Suppose we have a stable matrix
$$
\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}
$$
where $a + c < 0$ and $ac > bd$.  We want to know the probability
that, if we resample an element of this matrix, the result will be
stable.  By symmetry, there are two cases to consider; resampling 
$a$ is equivalent to resampling $d$ and resampling $b$ is equivalent
to resampling $c$.

Suppose that we resample $b$.  Since the trace does not involve $b$,
that condition will be satisfied automatically and we only need to
examine the determinant.  We will consider seperately the cases of $d
< 0$ and $d > 0$.  When $d > 0$, the condition becomes $ac/d > b$,
which is satisfiable when $ac > -d$ and results in probability $(ac/d
+ 1)/2$ of remaining stable.  When $d < 0$, the condition becomes
$ac/d < b$, which is satisfiable when $ac > d$, so we have probability
$(1 - ac/d)/2$.

Thus, the probability that a random stable matrix will remain stable
upon resampling $b$ is given as follows:
% \begin{strip}
\begin{align*}
&{\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,
  P(ac>b'd, a+c < 0) \over
\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,1} = \cr
&{{{1 \over 32} \int_0^1 dd \int_{d > ac > -d} da\,dc \int_{-1}^{ac/d} db \,
({ac \over d} + 1) + {1 \over 32} \int_{0}^{1} dd \int_{ac>d} da\,dc \int_{-1}^{1} db 2 \atop + 
{1 \over 32} \int_{-1}^0 dd \int_{-d > ac > d} da\,dc \int_{ac/d}^1 db\, (1
- {ac \over d}) + {1 \over 32} \int_{-1}^{0} dd \int_{ac>-d} da\,dc \int_{-1}^{1} db 2}
\over
{{{1 \over 16} \int_0^1 dd \int_{d > ac > -d} da\,dc \int_{-1}^{ac/d} db\,1
+ {1 \over 16} \int_{0}^{1} dd \int_{ac>d} da\,dc \int_{-1}^{1} db 1} \atop
{+ {1 \over 16} \int_{-1}^0 dd \int_{-d > ac > d} da\,dc \int_{ac/d}^1 db\,1} + {1 \over 16} \int_{-1}^{0} dd \int_{ac>-d} da\,dc \int_{-1}^{1} db 1}}
\end{align*}
% \end{strip}

In order to evaluate the integrals, we begin with the
integrals over $b$, which are particularly easy because none of the
integrands depends upon $b$.  Once we do that and make the change of
variables $d \mapsto -d$ in the second integrals in the numerator and
denominator, we can combine terms and simplify the expression to the
following:
$$
{\int_0^1 dd \int_{d > ac > -d} da\,dc ({ac \over d} + 1)^2 + \int_0^1 dd \int_{d > ac} da\,dc\, 4 \over
{2 \int_0^1 dd \int_{d > ac > -d} da\,dc ({ac \over d} + 1)} + 2 \int_0^1 dd \int_{d > ac} da\,dc\, 2}
$$
For $d > 0$ we have
$$
\int_{-d<ac<d} da\,dc \, f(a,c) = \int_{-1}^{-d} da \int_{d \over a}^{- d \over a} dc \, f(a,c) + \int_{-d}^{d} da \int_{-1}^{1} dc \, f(a,c) + \int_{d}^{1} da \int_{-d \over a}^{d \over a} dc \, f(a,c) 
%\over {\int_{-1}^{-d} da \int_{d \over a}^{- d \over a} dc + \int {-d}^{d} da \int_{-1}^{1} dc + \int_{d}^{1} da \int_{-d \over a}^{d \over a} dc}
$$
and 
$$
\int_{ac>d} da\,dc \, f(a,c) = \int_{-1}^{-d} da \int_{-1}^{d \over a} dc \, f(a,c) + \int_{d}^{1} da \int_{d \over a}^1 dc \, f(a,c)$$

% Thus, our expression becomes
% $$
% {\int_0^1 dd \int_{-1}^{-d} da \int_{-1}^{-d/a} dc\, ({ac \over d} + 1)^2 +
% \int_0^1 dd \int_{-d}^{d} da \int_{-1}^{1} dc\, ({ac \over d} + 1)^2 +
% \int_0^1 dd \int_d^1 da \int_{-d/a}^{1} dc\, ({ac \over d} + 1)^2
% \over
% 2\int_0^1 dd \int_{-1}^{-d} da \int_{-1}^{-d/a} dc\, ({ac \over d} + 1) +
% 2 \int_0^1 dd \int_{-d}^{d} da \int_{-1}^{1} dc\, ({ac \over d} + 1) +
% 2 \int_0^1 dd \int_d^1 da \int_{-d/a}^{1} dc\, ({ac \over d} + 1) }
% $$

Suppose that we resample $a$.  Then we have to satisfy the conditions
$a + c < 0$ and $ac > bd$ for fixed values of $b,c,d$.  We will
consider seperately the cases $c > 0$ and $c < 0$.  When $c > 0$, our
inequalities become $a < -c$ and $a > bd/c$.  For these to be
consistent, we must have $-c > bd/c$, or $bd < -c^2$ and, when we do,
we have a probability $(-c-bd/c)/2$ of reasmpling an $a$ that keeps
our matrix stable.  When $c < 0$, the conditions become $a < -c$ and
$a < bd/c$, which will have a solution as long as $bd/c > -1$, or $bd
< -c$.  Then the probability of resampling an $a$ which keeps the
matrix stable is $(\min(-c, bd/c) + 1)/2$.

Thus, the probability that a random stable matrix will remain stable upon resampling $a$ is given as follows:
\begin{align*}
&\frac{\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,
  P(a'c>bd, a'+c < 0)}{\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,1} = \\
&\frac{\genfrac{}{}{0pt}{}{{1 \over 32} \int_0^1 dc \int_{bd < -c^2} db\,dd \int_{\max(bd/c,-1)}^{-c} da \,
(-c - \max({bd \over c},-1)}{+
{1 \over 32} \int_{-1}^0 dc \int_{bd < -c} db\,dd \int_{-1}^{\min(-c,
bd/c)} da\, (\min(-c, bd/c) + 1)}}{{1 \over 16} \int_0^1 dc \int_{bd < -c^2} db\,dd \int_{bd/c}^{-c} da\,1
+
{1 \over 16} \int_{-1}^0 dc \int_{bd < -c} db\,dd \int_{-1}^{\min(-c, bd/c)} da\,1}
\end{align*}

For the case $c>0, {bd \over c} < a < -c, bd < -c^2$ we have two possibilities. For $bd < -c$ we have
$$
\int_{-1}^{-c} dd\, \int_{c/d}^1 db\, + \int_c^1 dd\, \int_{-1}^{-c/d} db,
$$
and for $bd > c$
$$
\int_{-1}^{-c} dd\, \int_{-c^2/d}^{-c/d} db\, + \int_{-c}^{-c^2} dd\, \int_{-c^2/d}^{1} db +
\int_{c^2}^{c} dd \int_{-1}^{-c^2/d} db +
\int_{c}^{1}dd\,\int_{-c/d}^{-c^2/d}db.
$$
For the case $c < 0, a < -c, a < bd/c$ we also have two possibilities. For $bd < -c^2$ we have
$$
\int_{-1}^{-c^2} dd\, \int_{-c^2/d}^1 db\, + \int_{c^2}^1 dd\, \int_{-1}^{-c^2/d} db,
$$
and for $bd > -c^2$
$$
\int_{-1}^{-c^2} dd\, \int_{-1}^{-c^2/d} db\, + \int_{-c^2}^{c^2} dd\, \int_{-1}^{1} db + 
\int_{c^2}^{1} dd \int_{-c^2/d}^{1} db.
$$

>>>>

\section{$2 \times 2$ systems}
We call a matrix stable when the real part of all its eigenvalues is negative.  For a $2 \times 2$ matrix, this amounts to the conditions $T < 0$ and $D >
0$ where $T$ and $D$ denote the trace and the determinant.

Suppose we have a stable matrix
$$
\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}
$$
where $a + c < 0$ and $ac > bd$. We will consider the case in which each of the parameters defining the matrix is sampled from the uniform distribution $\mathcal{U}(-1,1)$ so that the parameter space corresponds to the $d$-dimensional hypercube, $H^d$, of edge length $r=2$, centered about the origin, and having volume $r^d$.  We want to know the probability that, if we resample an element of this matrix, the result will be stable.  By symmetry, there are two cases to consider; resampling $a$ is equivalent to resampling $c$ and resampling $b$ is equivalent
to resampling $d$.

Suppose that we resample $b$.  We want to compute the probability that a random stable matrix will remain stable upon resampling $b$, which is
% \begin{strip}
\begin{align}\label{eq:condprob}
P\left(\begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right)
& = \frac{P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right)}.
\end{align}
The denominator of \ref{eq:condprob} is given by
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{ac>bd \atop a+c<0} \atop H^4} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprob}
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a' & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\begin{table}[h]
\begin{tabular}{ c || c | c }
\hline
matrix & \specialcell{probability of stability\\to perturbation} & probability of stability\\
\hline
  $\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}$ & 0.62 & 0.25 \\
  $\begin{pmatrix}
a & b \\
d & 0
\end{pmatrix}$, $\begin{pmatrix}
0 & b \\
d & c
\end{pmatrix}$ & 0.5 & 0.25 \\
  $\begin{pmatrix}
a & 0 \\
d & c
\end{pmatrix}$, $\begin{pmatrix}
a & b \\
0 & c
\end{pmatrix}$ & 0.67 & 0.25 \\
$\begin{pmatrix}
a & 0 \\
0 & c
\end{pmatrix}$ & 0.5 & 0.25 \\
\end{tabular}
\caption{Probability of matrix stability under resampling and \emph{a priori} stability. All matrices not listed have probability $0$ of stability.}\label{tab:structstabmat}
\end{table}

The probability of \emph{a priori} stability and of stability to perturbations for all $2 \times 2$-matrices are given in Table \ref{tab:structstabmat}.

%---------------------------------------------

% We will consider seperately the cases of $d
% < 0$ and $d > 0$.  When $d > 0$, the first condition becomes $ac/d > b$.
% Since $b > -1$, this can only be satisfied when $ac/d > -1$, i.e. when $ac > -d$.  In the subcase that $ac < d$, we then have the limits $-1 < b < ac/d$ and in the subcase $ac < d$, we have the limits $-1 < b < 1$.  When $d < 0$, the condition becomes
% $ac/d < b$, which is satisfiable when $ac > d$.  In the subcase where $ac > -d$, we have $ac/d < b < 1$ and, in the subcase where $ac < -d$, we have $-1 < b <
% 1$.

% Thus, the probability that a random stable matrix will remain stable
% upon resampling $b$ is given as follows:
% % \begin{strip}
% \begin{align*}
% P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)
% & = \frac{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable and } \begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right)}
% \end{align*}
% which becomes
% \begin{align*}
% &{\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,
%   P(ac>b'd, a+c < 0) \over
% \int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,1} = \cr
% &{{{1 \over 32} \int_0^1 dd \int_{d > ac > -d} da\,dc \int_{-1}^{ac/d} db \,
% ({ac \over d} + 1) + {1 \over 32} \int_{0}^{1} dd \int_{ac>d} da\,dc \int_{-1}^{1} db 2 \atop +
% {1 \over 32} \int_{-1}^0 dd \int_{-d > ac > d} da\,dc \int_{ac/d}^1 db\, (1
% - {ac \over d}) + {1 \over 32} \int_{-1}^{0} dd \int_{ac>-d} da\,dc \int_{-1}^{1} db 2}
% \over
% {{{1 \over 16} \int_0^1 dd \int_{d > ac > -d} da\,dc \int_{-1}^{ac/d} db\,1
% + {1 \over 16} \int_{0}^{1} dd \int_{ac>d} da\,dc \int_{-1}^{1} db 1} \atop
% {+ {1 \over 16} \int_{-1}^0 dd \int_{-d > ac > d} da\,dc \int_{ac/d}^1 db\,1} + {1 \over 16} \int_{-1}^{0} dd \int_{ac>-d} da\,dc \int_{-1}^{1} db 1}}
% \end{align*}
% % \end{strip}

% In order to evaluate the integrals, we begin with the
% integrals over $b$, which are particularly easy because none of the
% integrands depends upon $b$.  Once we do that and make the change of
% variables $d \mapsto -d$ in the second integrals in the numerator and
% denominator, we can combine terms and simplify the expression to the
% following:
% $$
% \frac{\int_0^1 dd \int_{ac > d} da\,dc\, 4 + \int_0^1 dd \int_{d > ac > -d} da\,dc ({ac \over d} + 1)^2}{2 \int_0^1 dd \int_{ac > d} da\,dc\, 2 + 2 \int_0^1 dd \int_{d > ac > -d} da\,dc ({ac \over d} + 1)}
% $$

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.5\columnwidth]{fig/2x2-resample-b.pdf}
% \caption{{\bf Resampling b.} (A) The region defined by $ac > d$ (B) The region defined by $d > ac > -d$}
% \label{fig:2x2-resample-b}
% \end{figure}

% For $ac > d$ we divide the $a$-axis into two parts indicated in Fig. \ref{fig:2x2-resample-b}A
% $$
% \int_{ac>d} da\,dc \, f(a,c) = \int_{-1}^{-d} da \int_{-1}^{d \over a} dc \, f(a,c) + \int_{d}^{1} da \int_{d \over a}^1 dc \, f(a,c)$$
% for the numerator $f(a,c) = 4$ and the integral is equal to 2. For $d > ac > -d$ we divide the $a$-axis into three parts indicated in Fig. \ref{fig:2x2-resample-b}B
% $$
% \int_{d > ac > -d} da\,dc \, f(a,c) = \int_{-1}^{-d} da \int_{d \over a}^{- d \over a} dc \, f(a,c) + \int_{-d}^{d} da \int_{-1}^{1} dc \, f(a,c) + \int_{d}^{1} da \int_{-d \over a}^{d \over a} dc \, f(a,c)
% % \over {\int_{-1}^{-d} da \int_{d \over a}^{- d \over a} dc + \int {-d}^{d} da \int_{-1}^{1} dc + \int_{d}^{1} da \int_{-d \over a}^{d \over a} dc}
% $$
% for the numerator $f(a,c) = (\frac{ac}{d} + 1)^2$ and the integral is equal to $\frac{32}{9}$ and for the denominator $f(a,c) = \frac{ac}{d} + 1$ and the integral is equal to $3$. Therefore
% $$P\left(\begin{pmatrix}
% a & b' \\
% d & c
% \end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
% a & b \\
% d & c
% \end{pmatrix} \textrm{stable } \right) = \frac{25}{36} \approx 0.69.$$

% % Thus, our expression becomes
% % $$
% % {\int_0^1 dd \int_{-1}^{-d} da \int_{-1}^{-d/a} dc\, ({ac \over d} + 1)^2 +
% % \int_0^1 dd \int_{-d}^{d} da \int_{-1}^{1} dc\, ({ac \over d} + 1)^2 +
% % \int_0^1 dd \int_d^1 da \int_{-d/a}^{1} dc\, ({ac \over d} + 1)^2
% % \over
% % 2\int_0^1 dd \int_{-1}^{-d} da \int_{-1}^{-d/a} dc\, ({ac \over d} + 1) +
% % 2 \int_0^1 dd \int_{-d}^{d} da \int_{-1}^{1} dc\, ({ac \over d} + 1) +
% % 2 \int_0^1 dd \int_d^1 da \int_{-d/a}^{1} dc\, ({ac \over d} + 1) }
% % $$

% Now suppose we resample $a$.  Then we have to satisfy the conditions
% $a + c < 0$, $ac > bd$, and $-1 < a < 1$ for fixed values of $b,c,d$.  We will
% consider seperately the cases $c > 0$ and $c < 0$.  When $c > 0$, our
% inequalities become $a < -c$ and $a > bd/c$.  For these to be
% consistent, we must have $-c > bd/c$, or $bd < -c^2$.  In the subcase where $bd < -c$, we have $-1 < a < -c$ and, in the subcase where $-c < bd < -c^2$, we have $bd/c < a < -c$. When $c < 0$, the conditions become $a < -c$ and
% $a < bd/c$, which will have a solution as long as $bd/c > -1$, or $bd
% < -c$.  In the subcase where $bd < -c^2$, we have $-1 < a < bd/c$ and, in the subcase where $bd > -c^2$, we have $-1 < a < -c$.

% Thus, the probability that a random stable matrix will remain stable upon resampling $a$ is given as follows:
% \begin{align*}
% &\frac{\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,
%   P(a'c>bd, a'+c < 0)}{\int_{ac>bd \atop a+c<0} da\,db\,dc\,dd\,1} = \\
% &\frac{\genfrac{}{}{0pt}{}{{1 \over 32} \int_0^1 dc \int_{bd < -c^2} db\,dd \int_{\max(bd/c,-1)}^{-c} da \,
% (-c - \max({bd \over c},-1)}{+
% {1 \over 32} \int_{-1}^0 dc \int_{bd < -c} db\,dd \int_{-1}^{\min(-c,
% bd/c)} da\, (\min(-c, bd/c) + 1)}}{{1 \over 16} \int_0^1 dc \int_{bd < -c^2} db\,dd \int_{bd/c}^{-c} da\,1
% +
% {1 \over 16} \int_{-1}^0 dc \int_{bd < -c} db\,dd \int_{-1}^{\min(-c, bd/c)} da\,1}
% \end{align*}

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.5\columnwidth]{fig/2x2-resample-a.pdf}
% \caption{{\bf Resampling a.} (A) The region defined by $bd < -c$ (B) The region defined by $-c < bd < -c^2$ (C) The region defined by $bd < -c^2$ (D) The region defined by $bd > -c^2$}
% \label{fig:2x2-resample-a}
% \end{figure}

% For the case $c>0, {bd \over c} < a < -c, bd < -c^2$ we have two possibilities. For $bd < -c$ we have
% $$
% \int_{-1}^{-c} dd\, \int_{c/d}^1 db\, + \int_c^1 dd\, \int_{-1}^{-c/d} db,
% $$
% and for $bd > c$
% $$
% \int_{-1}^{-c} dd\, \int_{-c^2/d}^{-c/d} db\, + \int_{-c}^{-c^2} dd\, \int_{-c^2/d}^{1} db +
% \int_{c^2}^{c} dd \int_{-1}^{-c^2/d} db +
% \int_{c}^{1}dd\,\int_{-c/d}^{-c^2/d}db.
% $$
% For the case $c < 0, a < -c, a < bd/c$ we also have two possibilities. For $bd < -c^2$ we have
% $$
% \int_{-1}^{-c^2} dd\, \int_{-c^2/d}^1 db\, + \int_{c^2}^1 dd\, \int_{-1}^{-c^2/d} db,
% $$
% and for $bd > -c^2$
% $$
% \int_{-1}^{-c^2} dd\, \int_{-1}^{-c^2/d} db\, + \int_{-c^2}^{c^2} dd\, \int_{-1}^{1} db +
% \int_{c^2}^{1} dd \int_{-c^2/d}^{1} db.
% $$

%---------------------------------------------
